Designing Qusecs Optimally: Quasi-uniform or Self-similar Layered Materials as Recursively Decomposed Solutions of Geometric Constraint Systems

### 2015-05-16 doc/GMP2015/revision_2/s03_drp.tex
Revision 2. 24 hours to submission.

\section{The canonical DR-plan}
\label{sec:DRP}
% In the problem of the optimal DR-Plan there is generally not a unique plan.
% % Indeed, we will prove a union of $N$ isostatic subgraphs will result in $N$ unique plans, but that at the $N^{\text{th}}$ level of the tree it will always be the same. Therefore, all choices of decomposition are in some sense equivalent. The theorem we seek to prove is thus:
% However, we will show that regardless of which children are chosen for the plan, so long as they satisfy the definition of an optimal DR-plan, the recombination will require solving of the same systems. Being the smallest such structure that offers this, the definition of an optimal DR-plan could be considered the canonical DR-plan.
% % To assist in showing this, we prove this core theorem throughout this section:

@@

The NP-hardness of the DR-planning problem for 2D bar-joint graphs is partly the consequence of possibly exponential number of DR-plans. On the other hand, although the complete DR-plan is unique it could have large average fan-in and exponentially many nodes making it far from optimal.

In this section, we define a \dfn{canonical} plan to capture those aspects of an optimal DR-plan that mimic a complete DR-plan's uniqueness, while for independent (underconstrained or isostatic graphs), we show that the nonunique parts do not affect optimality.

\sidenote{In this section, $G$ is always assumed to be isostatic.}

\begin{definition}\label{def:canonical_drplan}
    The \dfn{canonical DR-plan} satisfies the following three properties:
    (1) it is a DR-plan;
    (2) children are rigid vertex-maximal proper subgraphs of the parent; and
    (3) if all pairs of rigid vertex-maximal proper subgraphs intersect trivially then all of them are children, otherwise take as children exactly two that intersect non-trivially.
\end{definition}

This definition gives the canonical DR-plan a surprisingly strong Church-Rosser property.

\begin{theorem}
\label{theorem:canonical_exists_and_is_optimal}
\label{theorem:canonical_is_optimal}
\label{theorem:main}
    A canonical DR-plan exists and any canonical DR-plan is optimal.
\end{theorem}

@@

Begin with $\comdrp{G}$ of a rigid 2D bar-joint graph $G$, for all nodes $C$ with children $C_1,\ldots,C_N$ keep children nodes according to the following rules:
\begin{enumerate}
   \item If $C_i \cap C_j$ is trivial then keep all $C_1,\ldots,C_N$ as children.
   \item If $C_i \cap C_j$ is rigid then select any two out of $C_1,\ldots,C_N$ as children.
\end{enumerate}

This directly satisfies properties 2 and 3 of a canonical DR-plan (see definition \ref{def:canonical_drplan}), because all the nodes in $\comdrp{G}$ are rigid vertex-maximal proper subgraphs. However, we still have to show property 1 (that this makes a DR-plan).
For rule (1), because we are beginning with a complete DR-plan, if we keep all the children it is still a DR-plan. For rule (2), we know that the union must be rigid as well and it cannot be anything other than $C$, otherwise we have found a larger rigid vertex-maximal proper subgraph.

If we begin with an isostatic graph, rigid can be replaced with isostatic throughout the construction and remain valid. The rigid proper subgraphs of an isostatic graph must be isostatic themselves.

Now we must show that a canonical DR-plan is optimal. The proof is by induction on the level of the plan; we show that following these rules always guarantees minimum fan-in at this level and does not increase fan-in on following levels.

@@

\medskip\noindent
\vemph{Rule 1:}
% If some pair is intersecting trivially, then in fact all pairs intersect trivially, thus all must be children by defn of canonical drplan
%
First, we claim that if the intersection is trivial all $C_1,\ldots,C_N$ must be children of $C$ in an optimal DR-plan. We do this by showing that no subset of the children can union to form $C$, thereby requiring all of them be included; because this is the only choice, it is the minimum fan-in any DR-plan could have for this node.
% must be optimal for this node.

Take the strict subset $S\subsetneq \{1,\ldots,N\}$ such that $U=\bigcup_{i\in S}{C_i}$ which we \vemph{assume} is isostatic. If $U\neq C$, then we just found a larger proper subgraph and the children were not vertex-maximal to begin with. So, it must be that $U=C$.
\usestwod
However, since $C_i \cap C_j$ is trivial then for $k\notin S$ we know, by Lemma \ref{lemma:combined_lemma}, point \ref{lemma:uc_intersection_makes_all_uc}, $U\cap C_k$ must be one or more trivial subgraphs. By definition of a DR-plan $C_k=C\cap C_k$ and we know that $U=C$ so $C_k=U\cap C_k$. Thus, $C_k$ is both one or more trivial subgraphs and a isostatic subgraph of $C$. Thus, the assumption is wrong and $U$ cannot be isostatic. As $C$ is isostatic, this means no proper subset of $C_1,\ldots,C_N$ can union to form $C$.

Furthermore, since a canonical DR-plan uses proper rigid \vemph{vertex-maximal} subgraphs as children, and because their pairwise intersection is trivial, it follows that any such node has at most as many children as a DR-plan without this restriction, because the union of the children have to contain all edges of the parent. Therefore, this is the optimal choice in this case.

\medskip\noindent
\vemph{Rule 2:} Second, we claim that if the intersection is isostatic, the choice of any two children will result in an optimal DR-plan.

If $C_i \cap C_j$ is isostatic, then, by Observation \ref{lemma:union_intersection}, $C_i \cup C_j$ is also isostatic. This means that, by Lemma \ref{lemma:combined_lemma}, point \ref{lemma:wc_intersection_makes_all_wc}, any two children of $C$ will union to $C$ itself. Thus, any two children can be chosen to make a canonical DR-plan.
% are potential choices for the optimal DR-plan as they all create equal fan-in (exactly two) at this level.

\newcommand{\induceonc}[1]{Idc\left(C,#1\right)}
\renewcommand{\induceonc}[1]{#1}
\newcommand{\iunion}[1]{\induceonc{I\cup\bigcup_{k\in S_N\setminus\{#1\}}{R_k}}}

However, to guarantee that any two are the \vemph{optimal} choice, it must ensure minimum fan-in over all descendants. Take the set $S_N=\{1,\dots,N\}$, then we call $I=\bigcap_{k\in S_N}{C_k}$ and $R_k=C\setminus C_k$. Suppose we select $i$ and $j$, where $i\neq j$, as the children. For convenience, we will assume all subgraphs are induced subgraphs on $C$. We know that $C=\induceonc{I\cup\bigcup_{k\in S_N}{R_k}}$ and $C_i=\iunion{i}$. The isostatic vertex-maximal subgraphs of $C_i$ are $(\iunion{i,1}),\ldots,(\iunion{i,i-1}),(\iunion{i,i+1}),\ldots,(\iunion{i,N})$ all of which intersect on isostatic subgraphs. So any two of these are viable children for $C_i$.
% Since
% \[C_i=Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i\}}{R_k}\right)\]
% the children of $C_i$ will be
% \[Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i,m\}}{R_k}\right)\]
% and
% \[Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i,n\}}{R_k}\right)\]
% % $C_i=Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i\}}{R_k}\right)$
% % the children of this node will be
% % $Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i,m\}}{R_k}\right)$
% % and
% % $Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i,n\}}{R_k}\right)$
% for arbitrary $m$ and $n$, where $i,j,m,n$ do not equal each other. \todo{Prove these are valid children? Or is this obvious?}
This continues for $N-1$ levels total, always with fan-in of two (the minimum possible), at which point every descendant of $C$ is some $\induceonc{I\cup R_k}$ for $k\in S_N$, with every $k$ appearing at least once. Thus, regardless of the choice of $C_i,C_j$, then their children, etc., the DR-plan has fan-in of two for every node for the next $N$ levels, at which point the nodes contain the same subgraphs. Therefore, the choice is optimal.
%
% \medskip\noindent
% This proof then applies to itself recursively to show that the fan-in of the children will also be minimum.
%
% -- say this at top `proof is by induction on the level of the dr=plan'
\end{proof}

This proof of this theorem relied on the following crucial observation and lemma that are also used for other types of qusecs in the application sections (\ref{sec:bodypin} and \ref{sec:pinnedline}) of the paper.

\begin{observation*}\label{lemma:union_intersection}
If $F_i$ and $F_j$ are isostatic subgraphs of the same
isostatic graph $F$, \
then the following hold:
(1) $F_i\cup F_j$ is not trivial;
(2) $F_i\cup F_j$ is underconstrained if and only if $F_i\cap F_j$ is trivial;
(3) $F_i\cup F_j$ is isostatic if and only if $F_i\cap F_j$ is isostatic; and
(4) $F_i\cap F_j$ is not underconstrained.
\end{observation*}

Then the following key properties hold at the nodes of a canonical DR-plan.

\begin{lemma*}\label{lemma:combined_lemma}
Let $C$ be a node of a canonical DR-plan, with distinct children $C_i$ and $C_j$
Then
\begin{enumerate}
    \item\label{lemma:wc_intersection_is_C}
    $C_i\cup C_j$ is isostatic if and only if $C_i\cup C_j = C$.

@@

\begin{remark}
The first item in the above lemma generalizes to the union of any number of children, $C_1,\ldots,C_k$, resulting in the desirable property of \dfn{Cluster Minimality}, defined in \cite{hoffman2001decompositionI} holding for these canonical-optimal DR-plans.
\end{remark}

@@

\myexample
\textsl{[DR-Plan for self-similar structure]}
This section details the decomposition of the graph in Figure \ref{fig:c2c3ofk33s}. The graph $G$ has only two isostatic vertex-maximal subgraphs, $G$ without the outermost triangle composed of $K_{3,3}$ graphs (triangle $1$) and $G$ without the inner triangle $N$. These intersect on $G$ without triangle $1$ and $N$ which is clearly isostatic. As explained in the proof of Theorem \ref{theorem:main} since there are only 2 possible children, their intersection must be a node 2 levels below the parent. Just as expected, it is on the third level, as it is a child of both of $G$'s children. Furthermore, it intersects trivially with the edges that connect triangle $1$ and $N$ to $2$ and $N-1$, respectively; these necessarily intersect with triangle $1$ and $N$ themselves trivially (by Theorem \ref{theorem:main}, part 1) making all of these children. The triangles then decompose into their constituent $K_{3,3}$ graphs, which then decompose into 9 edges.

@@

\begin{figure*}\centering%
  %
  \includegraphics[width=0.3\linewidth]{img/epsfromtikz/demo_graph_candrp_seq}
  \caption{The sequential canonical DR-plan of $G_{\Delta}$ from Figure~\ref{fig:demo_graph:graph}, which is optimal (as explained in the proof of Theorem~\ref{theorem:algo_complexity}). The children of the triangle are omitted. Compare to to the typical canonical DR-plan shown in Figure~\ref{fig:demo_graph:candrp}. Also, note that the bottom-left node, the triangle $bce$, is the intersection of the 3 children of $G_{\Delta}$ in $ComDRP(G_{\Delta})$, shown in Figure~\ref{fig:demo_graph:comdrp}.}
  \label{fig:demo_graph:candrpseq}
\end{figure*}%

\begin{theorem}\label{theorem:algo_complexity}
    There exists an \candrpcomplexityv\ algorithm to find a canonical DR-plan for any 2D isostatic bar-joint graph.
\end{theorem}

\begin{proof}
The first step of the algorithm is finding the isostatic vertex-maximal proper subgraphs of the input isostatic graph $G$. Do this by first dropping arbitrary edge $e$ from the edge set and running the pebble game algorithm \cite{Jacobs:1997:PG} on this subgraph, which is $O(|V|^2)$. The output of this will be a list of component-candidates. Then run the Frontier algorithm \cite{hoffman2001decompositionII} \cite{lomonosov2004graph} ($O(|V|)$) on each candidate along with edge $e$, which will find the minimal subgraph $D$ that contains the candidate and $e$. If $D=G$, move the candidate (before adding $e$) to the true-component list and remove it from the candidate-component list. If $D\subsetneq G$, remove all candidates that are subgraphs of $D$ from the candidate-component list and add $D$ back to the list. Continue with the next candidate. When you exhaust the candidates, move to the next step.

Test the intersection of any two of the true-components. If the intersection is trivial, the entire list becomes the children. Recursively apply the algorithm to each child.

If the intersection is non-trivial (necessarily isostatic), immediately compute the sequential decomposition down to $I=\bigcap_{k\in S_N}{C_k}$. The root of this idea is discussed in detail in the proof of Theorem \ref{theorem:main}, and we use the notation here. We construct a DR-plan whose node set is the same as that of a canonical/optimal DR-plan. Begin with the parent $C$. Its children will be $C\setminus R_1$ and $R_1$ plus the nodes and edges adjacent to $R_1$ that lie in $I$. Recursively apply the algorithm to the child containing $R_1$. The children of $C\setminus R_1$ will be $C\setminus (R_1\cup R_2)$ and $R_2$ plus the nodes and edges adjacent to $R_2$ that lie in $I$. And so on. At $N$ levels down, the 2 children will be $I$ and $R_N$ plus the nodes and edges adjacent to $R_N$ that lie in $I$. Apply the algorithm recursively to both children. See Figure~\ref{fig:demo_graph:candrpseq} for an example.
% Although the nodes containing $R_1,\ldots,R_N$ are underconstrained, simply take the children to be the roots of the forest

The reason for this construction is that the result is a tree with no duplicate nodes. The leaves of this tree will exactly be the edges of this graph. Therefore, the number of nodes in the tree is $O(|E|)$ which, for isostatic input graphs, is $O(|V|)$.

Therefore, the complexity of the algorithm as a whole is $O(|V|^3)$.
%
% \note In the best case, this can be $O(V^2)$ if the graph is a 2-tree
\end{proof}

% \begin{proof}
% The first step of the algorithm is finding the isostatic vertex-maximal proper subgraphs of the input isostatic graph $G$. Do this by first dropping arbitrary edge $e$ from the edge set and running the pebble game algorithm \cite{Jacobs:1997:PG} on this subgraph, which is $O(|V|^2)$. The output of this will be a list of component-candidates. Then run the Frontier algorithm \cite{hoffman2001decompositionII} \cite{lomonosov2004graph} ($O(|V|)$) on each candidate along with edge $e$, which will find the minimal subgraph $D$ that contains the candidate and $e$. If $D=G$, move $D$ to the true-component list and remove the candidate from the candidate-component list. If $D\subsetneq G$, check the union of $D$ with the all items in the true-component list; if the union is isostatic, halt and output just these two components as children. Otherwise, remove all candidates that are subgraphs of $D$ from the candidate-component list and add $D$. Continue with the next candidate. If you exhaust the candidates, output the entire true-component list.

>>>>

\section{Main Result: canonical DR-plan, Optimality, and Algorithm}
\label{sec:DRP}
% In the problem of the optimal DR-Plan there is generally not a unique plan.
% % Indeed, we will prove a union of $N$ isostatic subgraphs will result in $N$ unique plans, but that at the $N^{\text{th}}$ level of the tree it will always be the same. Therefore, all choices of decomposition are in some sense equivalent. The theorem we seek to prove is thus:
% However, we will show that regardless of which children are chosen for the plan, so long as they satisfy the definition of an optimal DR-plan, the recombination will require solving of the same systems. Being the smallest such structure that offers this, the definition of an optimal DR-plan could be considered the canonical DR-plan.
% % To assist in showing this, we prove this core theorem throughout this section:

@@

\subsection{Canonical DR-plan}
In this section, we define a \dfn{canonical} plan to capture those aspects of an optimal DR-plan that mimic the  uniqueness of a complete DR-plan, and we show that the nonunique aspects do not affect optimality for independent (underconstrained or isostatic) graphs. Furthermore, we give an efficient \candrpcomplexity\ algorithm to find the canonical DR-plan of any independent graph. The definition is as follows:

\begin{definition}\label{def:canonical_drplan}
    The \dfn{canonical DR-plan} \addin{of a graph $G$} satisfies the following three properties:
    (1) it is a DR-plan \addin{of $G$};
    (2) children are rigid vertex-maximal proper subgraphs of the parent; and
    (3) if all pairs of rigid vertex-maximal proper subgraphs intersect trivially then all of them are children, otherwise exactly two that intersect non-trivially are children.
\end{definition}

In this section and in section~\ref{sec:recomb}, any reference to a graph $G$ is assumed to be isostatic (i.e.\ well-constrained or $(k,l)$-tight).
% \sidenote{In this section, $G$ is always assumed to be isostatic.}

% This definition gives the canonical DR-plan a surprisingly strong Church-Rosser property.

Definition~\ref{def:canonical_drplan} gives the canonical DR-plan a surprisingly strong Church-Rosser property, which is made explicit in Theorem~\ref{theorem:main}, the main result of this section.

\begin{theorem}
\label{theorem:canonical_exists_and_is_optimal}
\label{theorem:canonical_is_optimal}
\label{theorem:main}
    A canonical DR-plan exists \addin{for a graph $G$} and any canonical DR-plan is optimal \addin{if $G$ is independent.}
\end{theorem}

@@

Begin with $\comdrp{G}$ of a rigid 2D bar-joint graph $G$, for all nodes $C$ with children $C_1,\ldots,C_N$  retain children nodes according to the following rules:
\begin{enumerate}
   \item If $C_i \cap C_j$ is trivial then retain all $C_1,\ldots,C_N$ as children.
   \item If $C_i \cap C_j$ is rigid then select any two out of $C_1,\ldots,C_N$ as children.
\end{enumerate}

This directly satisfies Properties (2) and (3) of a canonical DR-plan (see Definition~\ref{def:canonical_drplan}), because all the nodes in $\comdrp{G}$ are rigid vertex-maximal proper subgraphs,  which we shorten to {\em clusters}.  To show Property (1) holds (that this constitutes a DR-plan):
for Case 1 above,  since we start with a complete DR-plan, if we preserve all the children it is still a DR-plan; for Case 2 above, we know that the union must be rigid as well and it cannot be anything other than $C$, otherwise we would have found a larger rigid proper subgraph of $C$, contradicting vertex-maximality.

Note that if we begin with an isostatic graph, ``rigid'' can be replaced with ``isostatic'' throughout the construction and preserve the above properties. The rigid proper subgraphs of an isostatic graph must be isostatic themselves.

Next we show that a canonical DR-plan is optimal.

First, note  that any  DR-plan \addin{$R$} without the Property (2) of a canonical DR-plan can always be modified (by introducing intermediate nodes) to satisfy Property (2) without  increasing the max fan-in, since any rigid proper subgraph of a graph $C$ (a child of node $C$ of the DR-plan \addin{$R$})
is the subgraph of some \addin{cluster} of $C$.
Thus without loss of generality, we can assume that \cutout{the} \addin{an} optimal DR-plan satisfies Property (2) of a canonical DR-plan.

\addin{The proof of optimality of a canonical DR-plan is by induction on its height.  The base case trivially holds for canonical DR-plans of height 0, i.e.\ for single edges. The induction hypothesis is that canonical DR-plans of height $t$ are optimal for the root node.}
\addin{For the induction step consider a canonical DR-plan $R$ of  height $t+1$ rooted at a node $C$.}
Notice that $R$ represents a canonical DR-plan $R(C)$ for the graphs $C$ corresponding to each of its  descendant nodes.
Thus, from the induction hypothesis, we know that the $R(C_i)$ is optimal for $C_i$.

Thus it is sufficient to demonstrate a set of nodes $S$  that must be present in any  DR-plan $R$ for $C$ that satisfies Property (2), including a known optimal one; and furthermore, for any such DR-plan $R$,  either (Claim 1) $S$ must be the set of children of $C$; or (Claim 2) for all the ancestors $A$ of $S$, $R$ has the minimum possible fan-in of 2.

We show the two claims below.
The first claim is that \addin{for a node $C$ whose clusters have trivial pairwise intersections}, any DR-plan \cutout{$R$} \addin{of $C$} that satisfies Property (2) must also satisfy Property (3) at $C$, i.e.\ the set of children $S$ of $C$ consists of all clusters of $C$.
\addin{Because this is the only choice, it is the minimum fan-in at $C$ for any DR-plan for $C$ with Property (2), including a known optimal one.}
The second claim shows that in the case of nodes $C$ whose rigid, vertex-maximal proper subgraphs have  non-trivial pairwise intersections, \addin{every canonical DR-plan of $C$ that uses any possible choice} of two such subgraphs of $C$ as children \cutout{ (minimum possible fan-in at $C$)} results in \addin{a minimum possible fan-in of 2 in the ancestor nodes $A$ leading to the}     {\em same maximal antichain $S$ of descendants $D$ of $C$}. \addin{The antichain is maximal in the partial order of rigid subgraphs of $C$ under containment.} I.e.\ $S$ satisfies the property that every proper vertex-maximal rigid subgraph of $C$ is a superset of some $D$ in $S$; \addin{this follows from properties of maximal antichains} that no element of $S$ is contained in the union of other elements of $S$; and the union of elements of $S$ is $C$. Thus any  DR-plan  \cutout{$R$} that satisfies Property (2) \addin{and hence contains two or more of the rigid vertex-maximal proper subgraphs of $C$ as children} must \addin{also} contain every element of $S$. The two claims complete the proof that every canonical
DR-plan is optimal.

@@

\medskip\noindent
\vemph{Claim 1:}
% If some pair is intersecting trivially, then in fact all pairs intersect trivially, thus all must be children by defn of canonical drplan
%
A set of clusters $C_1,\ldots,C_N$ whose pairwise intersection is trivial, must be children of $C$ in an optimal DR-plan.

We prove this claim  by showing that the union of no subset of the children can be $C$, thereby requiring all of them to be included as children.

% must be optimal for this node.

We prove by contradiction.
Assume to the contrary that the strict subset $S\subsetneq \{1,\ldots,N\}$ such that $U=\bigcup_{i\in S}{C_i}$ is isostatic. If $U\neq C$, then we found a larger proper subgraph contradicting vertex-maximality of the $C_i$. So, it must be that $U=C$.
\usestwod
However, since $C_i \cap C_j$ is trivial then for $k\notin S$ we know, by Lemma \ref{lemma:combined_lemma}, Item \ref{lemma:uc_intersection_makes_all_uc}, $U\cap C_k$ must be one or more trivial, i.e.\ disconnected vertices. By definition of a DR-plan, $C_k=C\cap C_k$ and we know that $U=C$ so $C_k=U\cap C_k$. Thus, $C_k$ is (i) a collection of disconnected vertices, and (ii) an isostatic subgraph of $C$, which is impossible. As $C$ is isostatic, this means the union of no proper subset of $C_1,\ldots,C_N$ is isostatic, nor is it equal to $C$, proving Claim 1.

Furthermore, since a canonical DR-plan has nodes with proper rigid \vemph{vertex-maximal} subgraphs as children, if, as in this case, their pairwise intersection is trivial, it follows that any node has at most as many children as a DR-plan without this restriction, because the union of the children must contain all edges of the parent. Therefore, the canonical DR-plan is the optimal choice in this case of trivial intersections.

\medskip\noindent
\vemph{Claim 2:}
If some pair in the set of child clusters $C_1,\ldots,C_N$ of $C$ has an isostatic (nontrivial) intersection, then choosing any two as children (minimum possible fan-in) will result in the same maximal antichain of descendants of $C$.

\addin{To prove Claim 2}, notice that if  $C_i \cap C_j$ is isostatic, then, by Observation \ref{lemma:union_intersection}, $C_i \cup C_j$ is also isostatic. This means that, by Lemma \ref{lemma:combined_lemma}, \addin{Point} \ref{lemma:wc_intersection_makes_all_wc}, the union of any two children of $C$ is $C$ itself. Thus, any two children can be chosen to make a canonical DR-plan and that is the minimum fan-in possible for a node of the DR-plan.
% are potential choices for the optimal DR-plan as they all create equal fan-in (exactly two) at this level.

\newcommand{\induceonc}[1]{Idc\left(C,#1\right)}
\renewcommand{\induceonc}[1]{#1}
\newcommand{\iunion}[1]{\induceonc{I\cup\bigcup_{k\in [N]\setminus\{#1\}}{R_k}}}

However, to guarantee that any two are the \vemph{optimal} choice, it must ensure minimum fan-in over all descendants leading up to a common maximal antichain $S$ of subgraphs.

To prove this holds, take the set $[N]=\{1,\dots,N\}$, and denote $I=\bigcap_{k\in [N]}{C_k}$ and $R_k=C\setminus C_k$. Suppose  $C_i$ and $C_j$, where $i\neq j$,  are  the children. For convenience, we will assume all subgraphs are induced subgraphs of $C$. We know that $C=\induceonc{I\cup\bigcup_{k\in [N]}{R_k}}$ and $C_i=\iunion{i}$. The isostatic vertex-maximal subgraphs of $C_i$ are $(\iunion{i,1}),\ldots,(\iunion{i,i-1}),(\iunion{i,i+1}),\ldots,(\iunion{i,N})$ all of whose pairwise intersections are isostatic subgraphs. So any two of these are viable children for $C_i$.
% Since
% \[C_i=Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i\}}{R_k}\right)\]
% the children of $C_i$ will be
% \[Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i,m\}}{R_k}\right)\]
% and
% \[Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i,n\}}{R_k}\right)\]
% % $C_i=Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i\}}{R_k}\right)$
% % the children of this node will be
% % $Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i,m\}}{R_k}\right)$
% % and
% % $Idc\left(C,I\cup\bigcup_{k\in S_N\setminus\{i,n\}}{R_k}\right)$
% for arbitrary $m$ and $n$, where $i,j,m,n$ do not equal each other. \todo{Prove these are valid children? Or is this obvious?}
This continues for $N-1$ levels, always with fan-in of two (the minimum possible), at which point every descendant of $C$ is some $\induceonc{I\cup R_k}$ for $k\in [N]$, with every $k$ appearing at least once. At the last level, there are exactly two rigid proper vertex-maximal subgraphs, and hence a unique choice of pair of children. Thus, regardless of the sequence of choices of $C_i$ and $C_j$, and of their descendants at each level, the DR-plan has the optimal fan-in of two for every node for  $N$ levels,  and the collection of last level nodes contain the same maximal antichain of subgraphs (for all choices).
%
% \medskip\noindent
% This proof then applies to itself recursively to show that the fan-in of the children will also be minimum.
%
% -- say this at top `proof is by induction on the level of the dr=plan'
\end{proof}

This proof of this theorem relies on the following crucial observation and lemma. These will be used again in the application sections (\ref{sec:bodypin} and \ref{sec:pinnedline}) of the paper, with modifications to work for other types of qusecs.

\begin{observation*}\label{lemma:union_intersection}
If $F_i$ and $F_j$ are non-empty isostatic graphs then the following hold: \\
(1) $F_i\cup F_j$ is not trivial;
(2) $F_i\cup F_j$ is underconstrained if and only if $F_i\cap F_j$ is trivial;
(3) $F_i\cup F_j$ is isostatic if and only if $F_i\cap F_j$ is isostatic; and
(4) $F_i\cap F_j$ is not underconstrained.
\end{observation*}

The following key properties hold at the nodes of a canonical DR-plan.

\begin{lemma*}\label{lemma:combined_lemma}
Let $C$ be an isostatic node of a canonical DR-plan, with distinct children $C_1,C_2,\ldots, C_m$. Assume $i\ne j$.
Then
\begin{enumerate}
    \item\label{lemma:wc_intersection_is_C}
    $C_i\cup C_j$ is isostatic if and only if $C_i\cup C_j = C$.

@@

\begin{remark}
The first item in the above lemma generalizes to the union of any number of children, $C_1,\ldots,C_k$, resulting in the desirable property of \dfn{cluster minimality} (defined in \cite{hoffman2001decompositionI} and in Section \ref{sec:prev}) holding for \cutout{these} canonical-optimal DR-plans.
\end{remark}

@@

\myexample
\textsl{[DR-Plan for self-similar structure]}
This example details the decomposition of the graph in Figure \ref{fig:c2c3ofk33s}. The graph $G$ has only two isostatic vertex-maximal subgraphs, $G$ without the outermost triangle composed of $K_{3,3}$ graphs (triangle $1$) and $G$ without the inner triangle $N$. These intersect on $G$ without triangle $1$ and $N$ which is clearly isostatic. As explained in the proof of Theorem \ref{theorem:main} since there are only 2 possible children, their intersection must be a node 2 levels below the parent. Just as expected, it is on the third level, as it is a child of both of $G$'s children. Furthermore, it intersects trivially with the edges that connect triangle $1$ and $N$ to $2$ and $N-1$, respectively; these necessarily intersect with triangle $1$ and $N$ themselves trivially (by Theorem \ref{theorem:main}, part 1) making all of these children. The triangles then decompose into their constituent $K_{3,3}$ graphs, which then decompose into 9 edges.

@@

% \begin{figure*}\centering%
%   %
%   \includegraphics[width=0.3\linewidth]{img/epsfromtikz/demo_graph_candrp_seq}
%   \caption{The sequential canonical DR-plan of $G_{\Delta}$ from Figure~\ref{fig:demo_graph:graph}, which is optimal (as explained in the proof of Theorem~\ref{theorem:algo_complexity}). The children of the triangle are omitted. Compare to to the typical canonical DR-plan shown in Figure~\ref{fig:demo_graph:candrp}. Also, note that the bottom-left node, the triangle $bce$, is the intersection of the 3 children of $G_{\Delta}$ in $ComDRP(G_{\Delta})$, shown in Figure~\ref{fig:demo_graph:comdrp}.}
%   \label{fig:demo_graph:candrpseq}
% \end{figure*}%
\begin{figure*}\centering%
  %
  \includegraphics[width=0.3\linewidth]{img/svg/3xc2c3_candrp_seq}
  \caption{The sequential canonical DR-plan of $G_{demo}$ from Figure~\ref{fig:demo_graph:graph}, which is optimal (as explained in the proof of Theorem~\ref{theorem:algo_complexity}). The children of the triangle are omitted. Compare to to the typical canonical DR-plan shown in Figure~\ref{fig:demo_graph:candrp}. Also, note that the bottom-left node, the triangle, is the intersection of the 3 children of $G_{demo}$ in $ComDRP(G_{demo})$, shown in Figure~\ref{fig:demo_graph:comdrp}.}
  \label{fig:demo_graph:candrpseq}
\end{figure*}%

\begin{theorem}\label{theorem:algo_complexity}
    There exists an \candrpcomplexityv\ algorithm to find a canonical DR-plan for \addin{an input graph}.
\end{theorem}

\begin{proof}
We first describe the algorithm.
The first step of the algorithm, which we call $CanDRP(G)$, is finding the isostatic vertex-maximal proper subgraphs (clusters) of the input isostatic graph $G$. One of many ways to do this is  by first dropping arbitrary edge $e$ from the edge set and running the pebble game algorithm \cite{Jacobs:1997:PG} on this subgraph, which is $O(|V|^2)$. The output of this will be a list of cluster-candidates. The next step is to isolate the set of true-clusters by running the \frontier\ algorithm \cite{hoffman2001decompositionII} \cite{lomonosov2004graph} ($O(|V|)$) on each candidate along with edge $e$, to find the minimal subgraph $D$ that contains the candidate and $e$. If $D=G$, the candidate is moved (before adding $e$) to the true-cluster list and removed from the candidate-cluster list. If $D\subsetneq G$, all candidates that are subgraphs of $D$ are removed from the candidate-cluster list and $D$ is added back to the list.  The next candidate is considered until all candidates are exhausted.

The next step is to check the intersection of any two of the true-components. If the intersection is trivial, the entire list becomes the children, and the algorithm is recursively applied to each child. If the intersection is non-trivial (necessarily isostatic), the  decomposition is computed down to $I=\bigcap_{k=1}^{N}{C_k}$. (This is discussed in detail in the proof of Theorem \ref{theorem:main}). We construct a DR-plan whose node set is that of a canonical/optimal DR-plan. Begin with the parent $C$. Its children are $S_1$ and $CanDRP(T_1)$, where $S_1=C\setminus R_1$ and $T_1$ is $R_1$, together with all incident edges. Observe that these edges are incident on vertices from $I$ and that $T_1$ is underconstrained so its DR-plan is a forest where each root becomes a child of $C$. The children of $S_1$ are $S_2$ and $CanDRP(T_2)$, where $S_2=C\setminus (R_1\cup R_2)$ and $T_2$ is $R_2$ together with all incident edges. The process iterates. At $N$ levels down, the children are  $S_N=I$ and $CanDRP(T_N)$, where $T_N$ is $R_N$ together with all incident edges. Apply the algorithm recursively to both children.
Thus we obtain an efficient method for the case of DR-plan nodes whose rigid proper vertex-maximal subgraphs have non-trivial intersections, avoiding recomputation of the isostatic components of $C$.
See Figure~\ref{fig:demo_graph:candrpseq} for an example.

To summarize the above step of the algorithm:
if the intersection is trivial, recursively apply $CanDRP$ to the entire list of true-components and set the resulting trees as the children. Else, the intersection is non-trivial, and the children of the node are $CanDRP(C_1)$ and $CanDRP(T)$, where $C_1$ is the first true-component of the parent $C$ (chosen arbitrarily, could be any child) and $T$ is the underconstrained graph formed by $C\setminus C_1$ together with all incident edges  and the associated nodes in $C_1$.

% Although the nodes containing $R_1,\ldots,R_N$ are underconstrained, simply take the children to be the roots of the forest

The DR-plan  output by this algorithm is a tree where each node is a distinct subgraph. The leaves of this tree are the edges of the starting graph $G$. Therefore, the number of nodes in the tree is $O(|E|)$ which, for isostatic input graphs, is $O(|V|)$ and the complexity of the algorithm is $O(|V|^3)$.
%
% \note In the best case, this can be $O(V^2)$ if the graph is a 2-tree
\end{proof}

% \begin{proof}
% The first step of the algorithm is finding the isostatic vertex-maximal proper subgraphs of the input isostatic graph $G$. Do this by first dropping arbitrary edge $e$ from the edge set and running the pebble game algorithm \cite{Jacobs:1997:PG} on this subgraph, which is $O(|V|^2)$. The output of this will be a list of component-candidates. Then run the \frontier\ algorithm \cite{hoffman2001decompositionII} \cite{lomonosov2004graph} ($O(|V|)$) on each candidate along with edge $e$, which will find the minimal subgraph $D$ that contains the candidate and $e$. If $D=G$, move $D$ to the true-component list and remove the candidate from the candidate-component list. If $D\subsetneq G$, check the union of $D$ with the all items in the true-component list; if the union is isostatic, halt and output just these two components as children. Otherwise, remove all candidates that are subgraphs of $D$ from the candidate-component list and add $D$. Continue with the next candidate. If you exhaust the candidates, output the entire true-component list.
